---
title: "NET scData after SoupX, plus *all* adult pancreas reference samples from Tosti et al. - Preprocessing, no SCT, downsampling of all samples to minimal total read count"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, message=FALSE}
library(Seurat)
library(ggplot2)

library(readxl)
library(knitr)
library(data.table)
library(cowplot)
library(scuttle)

library(pheatmap)

library(future)
plan("multicore", workers = 2)
options(future.globals.maxSize = 10*2**32)
#plan("sequential")

data_storage_folder = "../Data/Processed"
if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)

version_string = "SoupX_6best_and_all_Tosti_pancreas_preprocessing_downsampled_to_500_total_reads"

```

```{r}
reload_data = T
if(reload_data) {
  load(file.path("../Data/Processed/", "All_Filtered_scData_SoupX_integer.Rdata"))

  sel_cells = WhichCells(all_ge_data_filtered, expression = sample_id_final %in% c("SP084_007","SP084_018","SP084_019","SP084_021", "SP084_027", "SP084_028"))

  all_ge_data_filtered = subset(all_ge_data_filtered, cells = sel_cells)

  table(all_ge_data_filtered$sample_id_final)
}
```


```{r}
if(reload_data) {
  d = fread("../Data/Tosti_pancreas/Raw/exprMatrix.tsv.gz")
  d1 = Matrix(as.matrix(d[, 2:ncol(d)]), sparse=T)
  tmp = strsplit(d$gene,"\\|")
  rownames(d1) = unlist(sapply(tmp,`[`,1))
  rm(d)
  gc()
  
  meta = as.data.frame(fread("../Data/Tosti_pancreas/Raw/meta.tsv"))
  rownames(meta) = meta$Cell
  
  sc_adult_pancreas = CreateSeuratObject(d1, project = "Tosti_adult_pancreas", meta.data = meta)

  sc_adult_pancreas@meta.data$orig.ident = paste0(sc_adult_pancreas$patient_ID)
  sc_adult_pancreas <- RenameCells(sc_adult_pancreas, add.cell.id = "Tosti")
}
```

```{r}
if(reload_data) {
  genes_NET = rownames(all_ge_data_filtered@assays$RNA@counts)
  genes_Tosti = rownames(sc_adult_pancreas@assays$RNA@counts)

  genes_shared = intersect(genes_NET, genes_Tosti)

  meta_cols_NET = colnames(all_ge_data_filtered@meta.data)
  meta_cols_Tosti = colnames(sc_adult_pancreas@meta.data)

  shared_cols = intersect(meta_cols_NET, meta_cols_Tosti)

  sc_adult_pancreas@meta.data$DS = "Tosti"
  all_ge_data_filtered@meta.data$DS = "NET"

  for (cc in meta_cols_NET[!meta_cols_NET %in% shared_cols]) {
      sc_adult_pancreas = AddMetaData(sc_adult_pancreas, NA, col.name=cc)
  }

  for (cc in meta_cols_Tosti[!meta_cols_Tosti %in% shared_cols]) {
      all_ge_data_filtered = AddMetaData(all_ge_data_filtered, NA, col.name=cc)
  }

  combined_sc_data = merge(subset(all_ge_data_filtered, features = genes_shared), subset(sc_adult_pancreas, features = genes_shared))
}

```

We down-sample read counts by dividing by sample specific size factors estimated as the ratio of the sample median of cell total counts to a target total count. Cells that have less than 150 detected genes after downsampling will be excluded. 

```{r}
if(reload_data) {
  cs = colSums(combined_sc_data@assays$RNA@counts)
  cs_per_group = tapply(cs, combined_sc_data$orig.ident, median)
  boxplot(cs ~ combined_sc_data$orig.ident, ylab="Cell total read sums", las=2, xlab="", log="y", main="Before downsampling") 

target_cs = 500 # set some reasonable target read count 

size_factors = pmax(cs_per_group/target_cs,1) # min of size factors is 1 as we will not increase read counts. 


set.seed(20220531)
tmp = scuttle::downsampleMatrix(combined_sc_data@assays$RNA@counts, 1/size_factors[combined_sc_data$orig.ident])

cs2 = colSums(tmp)
boxplot(cs2 ~ combined_sc_data$orig.ident, ylab="Cell total read sums", las=2, xlab="", log="y", main="after downsampling")  

old_meta = combined_sc_data@meta.data
combined_sc_data = CreateSeuratObject(tmp, meta.data = old_meta[colnames(tmp),])
combined_sc_data$nCount_RNA = colSums(combined_sc_data@assays$RNA@counts)
combined_sc_data$nFeature_RNA = colSums(combined_sc_data@assays$RNA@counts>0)
combined_sc_data = subset(combined_sc_data, cells = WhichCells(combined_sc_data, expression = (nFeature_RNA > 150) ))
rm(tmp2, old_meta)
gc()
}

table(combined_sc_data$orig.ident)
```




# Final preprocessing

Since there are several independent batches of samples with partially very different outcomes in terms of processing quality and cell viability we will use data set integration (Stuart and Butler et al., Cell 2019, doi: 10.1016/j.cell.2019.05.031) as implemented in Seurat v3.


```{r}

split_data_file = paste0(version_string, "_split.Rdata")
if(reload_data) {
  sc.list <- SplitObject(combined_sc_data, split.by = "orig.ident")

  sc.list <- lapply(X = sc.list, FUN = function(x) {
      x <- NormalizeData(x)
      x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
  })

  sc.features <- SelectIntegrationFeatures(object.list = sc.list)
  sc.list <- lapply(X = sc.list, FUN = function(x) {
    x <- ScaleData(x, features = sc.features, verbose = FALSE)
    x <- RunPCA(x, features = sc.features, verbose = FALSE)
  })
  
  save(sc.list, sc.features, file=file.path(data_storage_folder, split_data_file ) )
} else {
  load(file.path(data_storage_folder, split_data_file))
}
```

# Software versions

```{r}
sessionInfo()
```

